Bacterial Microbiome and Obesity Meta-analysis Sequence Processing Pipeline

### 2016-03-21 code/sequenceProcessing.Rmd
Update sequence processing Rmd

---
title: "Bacterial Microbiome and Obesity Meta-analysis Sequence Processing Pipeline"
author: "Marc Sze"
date: "January 26, 2016"
output: 
  html_document:
    keep_md: yes
bibliography: metaAnalysis.obesity.microbiome.bib
---
#### **Preamble**  
Each study is brought up one at a time.  This recount starts from time of download all the way to generation of the final shared file used for the meta-analysis in the study.  You will also need the latest version of mothur, SRA Toolkit, Metaphlan2, Python 2.7, and Python 3+ installed on your system.  There are also a few reference files that you will need.  The first is the [SILVA reference](http://www.mothur.org/wiki/Silva_reference_files) database and the second is the [RDP database](http://www.mothur.org/wiki/RDP_reference_files) of which we used release 102 and version 9 respectively.  In general, we tried to keep close to the Schloss SOP for processsing sequence reads.  However, where this was not possible (mostly due to a high loss of reads resulting in a high loss of samples) we defaulted to the processing used by the published authors.  Where it was not possible to use a *de novo* approach (mostly due to high numbers of unique sequences) and the authors used a closed-reference approach we used the RDP database instead of greengenes.

<br>
<br>

@@

```{r, engine='bash', eval=FALSE}
/share/scratch/schloss/mothur/mothur data1.batch
```

If you thought this was difficult to get data just wait until later on in this pipeline to see just how crazy it can get trying to get the data to a useable point.
<br>
<br>

##### **Ross et al. 2015:**
This data set consisted of normal individuals from the Cameron Count Hispanic Cohort (CCHC).  The paper for this data set can be found [here](http://microbiomejournal.biomedcentral.com/articles/10.1186/s40168-015-0072-y).  The project accession number in the SRA is SRP053023.  We can download the entire data set again with the code mentioned above but with one small substitution at the ending.  Here we go......

@@

for sample2 in $(ls *fastq)  
do  
  /share/scratch/schloss/mothur/mothur "#fastq.info(fastq=$sample2)"  
  rm *logfile  
done

for TRIM in $(ls ERR*.fasta)  
do  
  QFILE=${TRIM//fasta/qual}  
  /share/scratch/schloss/mothur/mothur "#trim.seqs(fasta=$TRIM, qfile=$QFILE, maxambig=5, maxhomop=8, qwindowaverage=25, qwindowsize=50, processors=8)"  
done

@@

```{r, engine='bash', eval=FALSE}
cat \*trim.fasta > combined.trim.fasta  
/share/scratch/schloss/mothur/mothur "#unique.seqs(fasta=combined.trim.fasta)"

@@

After this little program is run we follow the mothur 454 SOP as written and can use the following command to complete the processing using the put together *Ross.batch* file and running mothur.  One final difference from the standardized mothur approach that was made was during the screen.seqs command after alignment a hard cutoff *minlength=200* and *maxlength=250* was used.

```{r, engine='bash', eval=FALSE}
/share/scratch/schloss/mothur/mothur Ross.batch
```

@@

##### **Goodrich et al. 2014:**  
This is a second larger twin data set that builds on the Turnbaugh et al. findings.  The paper can be found [here](http://www.ncbi.nlm.nih.gov/pmc/articles/PMC4255478/).  The individuals that participated in the study were part of the TwinsUK registry and is somewhat heavily skewed toward women.  The raw sequence data can be found on the SRA under ERP006339.  So using wget we can download the data with the following code:

@@

```{r, engine='sh', eval=FALSE}
#!/bin/sh  
# Unpack sra files  
# mkdir sra.complete  
for start in \$(ls \*sra)  
do  
  fastq-dump $start  
  mv $start sra.complete  
done

# Unpack fastq files  
mkdir fastq.complete  
mkdir logfile.complete  
for info in $(ls *fastq)  
do  
  /share/scratch/schloss/mothur/mothur "#fastq.info(fastq=$info)"  
  mv $info fastq.complete  
  mv *.logfile logfile.complete  
done

# Screen the fasta file that came from the fastq  
mkdir fasta.complete  
for next in $(ls *fasta)  
do  
  /share/scratch/schloss/mothur/mothur "#screen.seqs(fasta=$next, maxambig=0, maxlength=275, processors=8)"  
  mv $next fasta.complete # should store silva file in here and then move it to working directory  
  mv *.logfile logfile.complete  
done

mv ~/twinsUK/fasta.complete/silva.v4.fasta ~/twinsUK  
# Align the screened fasta file to V4 region of SILVA  
mkdir good.fasta.complete  
for sample2 in $(ls *good.fasta)  
do  
  /share/scratch/schloss/mothur/mothur "#align.seqs(fasta=$sample2, reference=silva.v4.fasta, processors=8)"  
  mv $sample2 good.fasta.complete  
  mv *logfile logfile.complete  
done

#Screen the aligned sequences to specific location  
mkdir align.complete  
for sample3 in $(ls *.align)  
do  
  /share/scratch/schloss/mothur/mothur "#screen.seqs(fasta=$sample3, start=1968, end=11546, maxhomop=8, processors=8)"  
  mv $sample3 align.complete  
done

# Filter the screened sequences to get a fasta file back  
mkdir align.good.complete  
for sample4 in $(ls *good.align)  
do  
  /share/scratch/schloss/mothur/mothur "#filter.seqs(fasta=$sample4, vertical=T, trump=., processors=8)"  
  mv $sample4 align.good.complete  
  mv *logfile logfile.complete  
done

# Make the combined fasta file  
mkdir filter.fasta.complete  
cat *filter.fasta > combined.fasta  
mv *filter.fasta filter.fasta.complete

# Make the combined group file  
grep -e '^>' combined.fasta > test.fasta  
sed 's/>//' test.fasta > test2.fasta  
sed 's/\..*$//' test2.fasta > test3.fasta  
paste -d"\t" test2.fasta test3.fasta > combined.groups

After this program was run I thought perhaps naively that I could simply run a set a batch file and be done with this data set.  Well not quite.  The sheer size and the way our server was configured meant that if I called much more than 130GB of RAM my job would run the risk of starting over regularly. So I did use the batch method but only upto *pre.cluster*.  After this the *chimera.uchime* step was just taking too long so I seperated everything again and did the *chimera.uchime* separately.  This also meant having to redo the unique seqs step.  The code for this is below:

@@

So now the biggest challenge that I have found so far is getting the data into a distance matrix not based on the closed-reference approach.  The file is just too big for cluster.split so we have had to use the closed-reference approach for this data set.

```{r, engine='bash', eval=FALSE}
/share/scratch/schloss/mothur/mothur Goodrich2.batch
```

@@

##### **Escobar et al. 2014:**
This data set consisted of normal individuals from Medellin, Colombia South America.  The paper for this data set can be found [here](http://bmcmicrobiol.biomedcentral.com/articles/10.1186/s12866-014-0311-6).  The project accession number in the ENA is ERP003466.  We can download the entire data set again with the code mentioned above but with a few substitutions versus the previous two studies.  Here we go......

@@

Because of this difference there are a few changes in this processing versus the standardized mothur 454 protocol.  First, instead of the sff.info() command we have to use the fastq.info() command.  Second, and perhaps more importantly, because we don't have the sff file we have to use the quality scores.  Unlike the Ross, et al study we used the mothur specified defaults of *qwindowaverage=35* and *qwindowsize=50*.  It should also be noted that we had *maxambig=0* and *maxhomop=8*.  Since the data come seperated already we do not need an oligo file and all we have to do is loop the mothur *trim.seqs* command. You can run a bash or shell program to get all of this done with either *#!/bin/sh* or *#!/bin/bash*.  So in a text editor:

@@

for sample2 in $(ls *fastq)  
do  
  /share/scratch/schloss/mothur/mothur "#fastq.info(fastq=$sample2)"  
  rm *logfile  
done
<br><br>
for TRIM in $(ls SRR*.fasta)  
do  
  QFILE=\${TRIM//fasta/qual}  
/share/scratch/schloss/mothur/mothur "#trim.seqs(fasta=$TRIM, qfile=$QFILE, maxambig=0, maxhomop=8, qwindowaverage=35, qwindowsize=50, processors=8)"  
done

@@

/share/scratch/schloss/mothur/mothur "#unique.seqs(fasta=combined.trim.fasta)"

@@

```{r, engine='bash', eval=FALSE}
/share/scratch/schloss/mothur/mothur Escobar.batch
```

@@

Now this data set included the sff files so we can use sffinfo function and start from the very start of the mothur 454 pipleine.  Plus one for this data set so far.  The below program was used and you can use either *#!/bin/sh* or *#!/bin/bash* and text editor to create it and then run it.  Also you should be aware that this part of the processing will take a LONG time to get through.

@@

# Make a directory to store the reference files
mkdir reference.files
mv silva.bacteria.fasta reference.files
mv trainset9_032012.pds.tax reference.files
mv trainset9_032012.pds.fasta reference.files

@@

for sample2 in $(ls *fastq)
do
	/share/scratch/schloss/mothur/mothur "#fastq.info(fastq=$sample2)"	
	mv $sample2 completed.fastq
	mv *logfile logfile.completed

@@

# Trim the fastas according to quality 
mkdir first.trim.completed
for Trimming in $(ls *fasta)
do
	Quality=${Trimming//fasta/qual}
	/share/scratch/schloss/mothur/mothur "#trim.seqs(fasta=$Trimming, qfile=$Quality, maxambig=0, maxhomop=8, qwindowaverage=35, qwindowsize=50)"
	mv $Trimming first.trim.completed
	mv $Quality first.trim.completed
	mv *logfile logfile.completed
done

@@

```{r, engine='bash', eval=FALSE}
/share/scratch/schloss/mothur/mothur Zupancic.batch
```

With that you should now be able to generate the subsample shared file that was used for the downstream analysis.  This was a headache of a data set to work with.  I think the next one will be better. At least I hope so....
<br><br>

##### **Yatsunenko et al. 2012:**  
This data set was another famous one looking at the microbiome of indivdiuals at different geographical locations and pretty different diets.  The link to the paper is [here](http://www.ncbi.nlm.nih.gov/pmc/articles/PMC3376388/).  A couple problems from a data processing and sequence clean up point of view is that they used a HiSeq, single end, 100bp total length, sequencing run.  This wreaked absolute havoc with a lot of the downstream steps mostly with trying to figure out a way to reduce the artificial diversity within the files.  Oh... and the data was not on a public database but stored on MG-RAST.  I had a lot trouble downloading the data on command line all at once (maybe I'm just not smart enough to do it).  So I had to manually download the sequence files.Thats right I am just a glutton for punishment.  After you have downloaded the files the first part of the code is:

```{r, engine='bash', eval=FALSE}
#!/bin/sh  
mkdir fna.done  
mkdir logfiles.done  
mkdir bad.accnos.done

# Algin the sequences
mkdir good.fna.done  
for sample in $(ls *.fna)  
do  
  /share/scratch/schloss/mothur/mothur "#align.seqs(fasta=\$sample, reference=silva.bacteria.fasta, processors=8)"  
  mv $sample good.fna.done  
  mv *logfile logfiles.done  
done

For the next code it might be best to just double check that the start and end are what is written here.  To make it go faster I made a custom pcr.seqs based on the start and end of the sequences on the aligned file.  It shouldn't make a difference but.....you never know.

```{r, engine='bash', eval=FALSE}
#!/bin/sh

# Screen the sequences to the same start and end
mkdir align.done  
for sample2 in $(ls *.align)  
do  
  /share/scratch/schloss/mothur/mothur "#screen.seqs(fasta=$sample2, start=8, end=2448, maxhomop=8, processors=8)"  
  mv $sample2 align.done  
  mv *logfile logfiles.done  
done

# Filter the sequences
mkdir good.align.done  
for sample3 in \$(ls \*good.align)  
do  
  /share/scratch/schloss/mothur/mothur "#filter.seqs(fasta=\$sample3, vertical=T, trump=., processors=8)"  
  mv \$sample3 good.align.done  
  mv \*logfile logfiles.done  
done

# Create a groups file and combine the fasta and groups together
mkdir completed.good.groups  
for sample in \$(ls \*.good.filter.fasta)  
do  
  grep -e '^>' \$sample > test.fasta  
  sed 's/>//' test.fasta > test2.fasta  
  sed 's/_.*//' test2.fasta > test3.fasta  
  paste -d"\t" test2.fasta test3.fasta > \${sample}.groups

  rm test.fasta  
  rm test2.fasta  
  rm test3.fasta  
  mv ~/Yatsunenko.Nature/${sample}.groups  ~/Yatsunenko.Nature/completed.good.groups

done
```

From here I thought I might be able to get away with combining the data set and running everything together.  Well that was not the case since it is just too large for the memory I have readily available without having the job restarting.  The plan is to seperate everything back out again and run the *chimera.uchime* function.  Then combine them and run the *pre.cluster* function.  The code below shows what was used for this part:

```{r, engine='bash', eval=FALSE}
#!/bin/sh  
cat *fasta > combined.fasta  
cat *groups > combined.groups

/share/scratch/schloss/mothur/mothur "#unique.seqs(fasta=combined.fasta)"    
/share/scratch/schloss/mothur/mothur Yatsunenko.batch

done
```

##### **Human Microbiome Project (HMP) 2011:**  
This data set consisted of healthy individuals that were sampled at multiple body sites and underwent extensive medical work up for metadata collection.  In order to be included in this study you could not have any underlying medical conditions.  The project website can be found [here](http://hmpdacc.org/).  The project accession number in the SRA is SRP002860.  The sequence processing pipeline can be found in an ipython notebook located [here](http://nbviewer.jupyter.org/gist/pschloss/9815766/notebook.ipynb).  The one exception to the workflow provided in the notebook is that you will need to make sure that you do not use a closed reference data base approach for OTU clustering and instead use an open reference approach.
<br><br>

##### **Nam, et al. 2011:**  
This study recruited 20 individuals that were from the area surronding Kyung Hee University in South Korea and can be found [here](http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0022109).  The data from this project can be found under the accession number PRJDA6057 and we downloaded the data from DRR000776.  There is some small difference to the wget, which is mostly associated with the directory where the data is stored.  From here you can run the SRA tool kit command to get the sff file.  Once this is done you can then run mothur with the batch file and get the sub sample shared file that was used in this analysis.  We have also provided the oligo file in this repository *korean6.oligos* that was used so that you can easily use it on this data set.

```{r, engine='bash', eval=FALSE}
wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByRun/sra/DRR/DRR000/DRR000776

sff-dump DRR000776.sra

/share/scratch/schloss/mothur/mothur Nam.batch''

Compared to the previous data sets this was one of the most straight forward to actually process.
<br><br>

@@

# Trim the fastas according to quality 
mkdir first.trim.completed
for Trimming in $(ls *fasta)
do
	Quality=${Trimming//fasta/qual}
	/share/scratch/schloss/mothur/mothur "#trim.seqs(fasta=$Trimming, qfile=$Quality, maxambig=0, maxhomop=8, qwindowaverage=35, qwindowsize=50)"
	mv $Trimming first.trim.completed
	mv $Quality first.trim.completed
	mv *logfile logfile.completed
done

mkdir scrap.fasta.completed
mkdir scrap.qual.completed
mkdir trim.qual.completed
mv *scrap.qual scrap.qual.completed
mv *scrap.fasta scrap.fasta.completed
mv *trim.qual trim.qual.completed

@@

So for the next part I needed to make a group file so I renamed the fasta file to represent the sampleID label so that I could with more ease create the labels I needed within the group file. The code for it is....

@@

groupN="SRR327637 SRR327638 SRR327639 SRR327640 SRR327641 SRR327642 SRR327643 SRR327644 SRR327645 SRR327646 SRR327647 SRR327648 SRR327649 SRR327650 SRR327651 SRR327652 SRR327653 SRR327654 SRR327655 SRR327656 SRR327657 SRR327658 SRR327659 SRR327660 SRR327661 SRR327662 SRR327663 SRR327664 SRR327665 SRR327666 SRR327667 SRR327668 SRR327669 SRR327670 SRR327671 SRR327672 SRR327673 SRR327674 SRR327675 SRR327676 SRR327677 SRR327678 SRR327679 SRR327680 SRR327681 SRR327682 SRR327683 SRR327684 SRR327685 SRR327686 SRR327687 SRR327688 SRR327689 SRR327690 SRR327691 SRR327692 SRR327693 SRR327694 SRR327695 SRR327696 SRR327697 SRR327698 SRR327699 SRR327700 SRR327701"

for V in ${groupN}
do
	grep -e '^>' ${V}.trim.fasta > test.fasta
	sed 's/>//' test.fasta > test2.fasta
	sed "s/$/\t$V/" test2.fasta > ${V}.groups
	rm test.fasta
	rm test2.fasta
done

After you have done this you can run a batch file with mothur and that should generate the subsample shared file that we used for this anlaysis.

```{r, engine='bash', eval=FALSE}
/share/scratch/schloss/mothur/mothur Wu.batch
```
<br><br>

##### **Arumugam, et al. 2010 (MetaHit):**  
We only used the Danish individuals for this study and left the others alone since many had IBD or other disease conditions.  For this data set Metaphlan2 was used to generate the abundance data.  Metaphlan2 and how well it associates with 16S data is also discussed in the supplemental of the draft manuscript we are writing.  To test this we downloaded ~40 samples from the HMP data set which had both metagenome and 16S data.  We ran the same code that we used with the MetaHit data to generate the abundance data using Metaphlan2 and compared that result with the obtained 16S values.  So here is code that was used below:

```{r, engine='bash', eval=FALSE}
#!/bin/sh

PATH=\$PATH:~/metaphlan2  
export mpa_dir=~/metaphlan2  
mkdir completed.combined.fq

test_samples="MH0001 MH0002 MH0003 MH0004 MH0005 MH0006 MH0007 MH0008 
MH0009 MH0010 MH0011 MH0012 MH0013 MH0014 MH0015 MH0016 MH0017 MH0018 
MH0019 MH0020 MH0021 MH0022 MH0023 MH0024 MH0025 MH0026 MH0027 MH0028 
MH0030 MH0031 MH0032 MH0033 MH0034 MH0035 MH0036 MH0037 MH0038 MH0039 
MH0040 MH0041 MH0042 MH0043 MH0044 MH0045 MH0046 MH0047 MH0048 MH0049 
MH0050 MH0051 MH0052 MH0053 MH0054 MH0055 MH0056 MH0057 MH0058 MH0059 
MH0060 MH0061 MH0062 MH0063 MH0064 MH0065 MH0066 MH0067 MH0068 MH0069 
MH0070 MH0071 MH0072 MH0073 MH0074 MH0075 MH0076 MH0077 MH0078 MH0079 
MH0080 MH0081 MH0082 MH0083 MH0084 MH0085 MH0086"

for s in ${test_samples}  
do  
  cat ${s}*.1.fq > ${s}.combined.1.fq  
  cat ${s}*.2.fq > ${s}.combined.2.fq  
  mv ${s}*combined.1.fq completed.combined.fq  
  mv ${s}*combined.2.fq completed.combined.fq  
done

This code was used to combine the fastq files since there are multiple files per sample.  To actually implement the metaphlan2 program for these files see below:

```{r, engine='bash', eval=FALSE}
for s in ${test_samples}  
do  
  metaphlan2.py /mnt/EXT/Schloss-data/msze/test/danish.obesity.metagenome/raw.data.sep/completed.combined.fq/${s}.combined.1.fq,/mnt/EXT/Schloss-data/msze/test/danish.obesity.metagenome/raw.data.sep/completed.combined.fq/${s}.combined.2.fq --mpa_pkl db_v20/mpa_v20_m200.pkl --bowtie2db db_v20/mpa_v20_m200 --input_type fastq --bowtie2out profiled_samples/${s}.bowtie2.bz2 -o profiled_samples/${s}.txt

done
```

In order to combine all the files into a single data table the following code was run:
```{r, engine='bash', eval=FALSE}
python ~/metaphlan2/utils/merge_metaphlan_tables.py profiled_samples*.txt > profiled_samples/output/merged_abundance_table.txt
```

For the 16S and metaphlan2 comparison the following code was run to obtain the necessary data tables:

```{r, engine='bash', eval=FALSE}
#!/bin/sh  
PATH=\$PATH:~/metaphlan2  
export mpa_dir=~/metaphlan2

test_samples="SRS012849 SRS012969 SRS013098 SRS013216 SRS013638 SRS013639 SRS016437 SRS016438 SRS016990 SRS017916 SRS020622 SRS021153 SRS021219 SRS022093 SRS024663 SRS043667 SRS045528 SRS045739 SRS049402 SRS049896 SRS050026 SRS053356 SRS053573 SRS053649 SRS054352 SRS055533 SRS056273 SRS058070 SRS063489 SRS063518 SRS064973 SRS074670 SRS074964 SRS075078 SRS075341 SRS075963 SRS076929 SRS077086 SRS077194 SRS077294 SRS077335 SRS077502 SRS077552 SRS077753 SRS077849 SRS078242 SRS078419 SRS078665 SRS097889 SRS098514 SRS098571 SRS098644 SRS098717 SRS098827 SRS0100021 SRS101376 SRS101433 SRS103987 SRS0104197 SRS104311 SRS104400 SRS104485 SRS105153 SRS140492 SRS140513 SRS140645 SRS142503 SRS142505 SRS142599 SRS142712 SRS142890 SRS143070 SRS143181 SRS143342 SRS143417 SRS143598 SRS143780 SRS143876 SRS143991 SRS144362 SRS144506 SRS144537 SRS145497 SRS146764 SRS146812 SRS146813 SRS147022 SRS147088 SRS147139 SRS147346 SRS147445 SRS147652 SRS147766 SRS147919 SRS148196 SRS148424 SRS148721"

for s in ${test_samples}  
do  
#Download and unpack necessary files  
  wget -r -np -nd -k ftp://public-ftp.hmpdacc.org/Illumina/PHASEII/stool/${s}.tar.bz2  
tar xvjf ${s}.tar.bz2  
  mv ${s}/${s}.denovo_duplicates_marked.trimmed.1.fastq ~/metaphlan2  
  mv ${s}/${s}.denovo_duplicates_marked.trimmed.2.fastq ~/metaphlan2  
  metaphlan2.py ${s}.denovo_duplicates_marked.trimmed.1.fastq,${s}.denovo_duplicates_marked.trimmed.2.fastq --mpa_pkl db_v20/mpa_v20_m200.pkl --bowtie2db db_v20/mpa_v20_m200 --input_type fastq --bowtie2out profiled_samples/${s}.bowtie2.bz2 -o profiled_samples/${s}.txt  
  rm -r ${s}  
  rm ${s}.denovo_duplicates_marked.trimmed.1.fastq  
  rm ${s}.denovo_duplicates_marked.trimmed.2.fastq  
done
```

You can then go into the directory with all the seperate data tables and combined them into one with the following command:

```{r, engine='bash', eval=FALSE}
python ~/metaphlan2/utils/merge_metaphlan_tables.py profiled_samples*.txt > profiled_samples/output/merged_abundance_table.txt
```

Now I decided, perhaps foolishly, to manually curate the combined file.  What I mean by this is that I manually deleted viruses and eukaryotic cells that were identified instead of using a computer to do this.  I also then manually created data tables with phyla and the lowest taxonomic identification.  The data table with the lowest taxonomic identification information was used to generate alpha diversity measures and to complete and PERMANOVA analysis that was used.  These manually curated files are provided in this repository for you to use as you want.  Finally, I used relative abundance instead of subsampling in these samples.  So I had to redo the relative abundance after removing the viruses and eukaryotic cells.  This part was done in R.
<br><br>

##### **Turnbaugh, et al. 2008:**  
This is the very famous twin study of the microbiome and obesity.  The other two re-analysis articles [@finucane_taxonomic_2014; @walters_meta-analyses_2014] also use this data set.  The raw data is not avialable on the SRA but you can download a fasta file for either the V2 or V6 16S rRNA gene [here](https://gordonlab.wustl.edu/NatureTwins_2008/TurnbaughNature_11_30_08.html).  We used the V2 region.  For the most part it was relatively straight forward, just be aware that this took a lot of RAM to create a distance matrix.  The code I used is below:

```{r, engine='bash', eval=FALSE}
grep -e '^>' combined.trim.fasta > test.fasta  
sed 's/>//' test.fasta > test2.fasta    
sed 's/\s.*$//' test2.fasta > test3.fasta  
paste -d" " test2.fasta test3.fasta > V2.groups''
```

Once we have a groups file we can just create a batch file and put it into mothur to run and come back when it is completed.
```{r, engine='bash', eval=FALSE}
/share/scratch/schloss/mothur/mothur Turnbaugh.batch
```

>>>>

---
title: "Bacterial Microbiome and Obesity Meta-analysis Sequence Processing Pipeline"
author: "Marc Sze"
date: "January 26, 2016"
output: 
  html_document:
    keep_md: yes
bibliography: ../metaAnalysis.obesity.microbiome.bib
---
#### **Preamble**  
Each study is brought up one at a time.  This recount starts from time of download all the way to generation of the final shared file used for the meta-analysis in the study.  You will also need the latest version of mothur, SRA Toolkit, and Python 3+ installed on your system.  There are also a few reference files that you will need.  The first is the [SILVA reference](http://www.mothur.org/wiki/Silva_reference_files) database and the second is the [RDP database](http://www.mothur.org/wiki/RDP_reference_files) of which we used release 102 and version 9 respectively.  In general, we tried to keep as close as possible to the Schloss SOP for processsing sequence reads for either 454 or the MiSeq.  Most of the differences between the SOP as written and what was performed involved the use of flowgrams.  Most of the data in the SRA were stored in the Fastq format and this step was not possible.
<br>
<br>

@@

```{r, engine='bash', eval=FALSE}
mnt/EXT/Schloss-data/bin/mothur Baxter.batch
```

##### **Ross et al. 2015:**
This data set consisted of normal individuals from the Cameron Count Hispanic Cohort (CCHC).  The paper for this data set can be found [here](http://microbiomejournal.biomedcentral.com/articles/10.1186/s40168-015-0072-y).  The project accession number in the SRA is SRP053023.  We can download the entire data set again with the code mentioned above but with one small substitution at the ending.

@@

for sample2 in $(ls *fastq)  
do  
  mnt/EXT/Schloss-data/bin/mothur "#fastq.info(fastq=$sample2)"  
  rm *logfile  
done

for TRIM in $(ls ERR*.fasta)  
do  
  QFILE=${TRIM//fasta/qual}  
  mnt/EXT/Schloss-data/bin/mothur "#trim.seqs(fasta=$TRIM, qfile=$QFILE, maxambig=5, maxhomop=8, qwindowaverage=25, qwindowsize=50, processors=8)"  
done

@@

```{r, engine='bash', eval=FALSE}
cat \*trim.fasta > combined.trim.fasta  
mnt/EXT/Schloss-data/bin/mothur "#unique.seqs(fasta=combined.trim.fasta)"

@@

After this little program is run we follow the mothur 454 SOP pretty much as  written and can use the following command to complete the processing using the put together *Ross.batch* file and running mothur.  Moving forward for the screen.seqs command we use a minlength cutoff of 200 in addition to what is written in the 454 SOP for all data sets.

```{r, engine='bash', eval=FALSE}
mnt/EXT/Schloss-data/bin/mothur Ross.batch
```

@@

##### **Goodrich et al. 2014:**  
This is a second larger twin data set and the paper can be found [here](http://www.ncbi.nlm.nih.gov/pmc/articles/PMC4255478/).  The individuals that participated in the study were part of the TwinsUK registry and is somewhat heavily skewed toward women.  The raw sequence data can be found on the SRA under ERP006339.  So using wget we can download the data with the following code:

@@

```{r, engine='sh', eval=FALSE}
#!/bin/sh  
# Unpack sra files  
# mkdir sra.complete  
for start in $(ls *sra)  
do  
  fastq-dump $start  
  mv $start sra.complete  
done

#Unpack fastq files  
mkdir fastq.complete  
mkdir logfile.complete  
for info in $(ls *fastq)  
do  
	mnt/EXT/Schloss-data/bin/mothur "#fastq.info(fastq=$info)"  
	mv $info fastq.complete  
	mv *.logfile logfile.complete  
done

#Trim the fasta file that came from the fastq  
mkdir fasta.trimming.complete  
for next in $(ls *fasta)  
do  
	QualFile=${next//fasta/qual}
	/mnt/EXT/Schloss-data/bin/mothur "#trim.seqs(fasta=$next, qfile=$QualFile, maxambig=0, maxhomop=8, qwindowaverage=35, qwindowsize=50)"  
	mv $next fasta.trimming.complete  
	mv $QualFile fasta.trimming.complete
	mv *.logfile logfile.complete  
done

#Create a groups file for each sample
for sample2 in $(ls *trim.fasta)
do
	groupfile=${sample2//fasta/groups}
	grep -e '^>' $sample2 > test.fasta  
	sed 's/>//' test.fasta > test2.fasta  
	sed 's/\..*$//' test2.fasta > test3.fasta  
	paste -d"\t" test2.fasta test3.fasta > ${groupfile}

#Make the combined fasta file  
mkdir combined.complete  
cat *trim.fasta > combined.fasta  
mv *trim.fasta combined.complete
cat *trim.groups > combined.groups
mv *trim.groups combined.complete
```

From here you can run a batch file with mothur to get the final output files.

@@

@@

##### **Escobar et al. 2014:**
This data set consisted of normal individuals from Medellin, Colombia South America.  The paper for this data set can be found [here](http://bmcmicrobiol.biomedcentral.com/articles/10.1186/s12866-014-0311-6).  The project accession number in the ENA is ERP003466.  We can download the entire data set again with the code mentioned above but with a few substitutions versus the previous two studies.

@@

Because of this difference there are a few changes in this processing versus the standardized mothur 454 protocol.  First, instead of the sff.info() command we have to use the fastq.info() command.  Second, and perhaps more importantly, because we don't have the sff file we have to use the quality scores.  Unlike the Ross, et al study we used the mothur specified defaults of *qwindowaverage=35* and *qwindowsize=50*.  Since the data come seperated already we do not need an oligo file and all we have to do is loop the mothur *trim.seqs* command. You can run a bash or shell program to get all of this done with either *#!/bin/sh* or *#!/bin/bash*.  So in a text editor:

@@

for sample2 in $(ls *fastq)  
do  
  mnt/EXT/Schloss-data/bin/mothur "#fastq.info(fastq=$sample2)"  
  rm *logfile  
done
<br><br>
for TRIM in $(ls SRR*.fasta)  
do  
  QFILE=\${TRIM//fasta/qual}  
mnt/EXT/Schloss-data/bin/mothur "#trim.seqs(fasta=$TRIM, qfile=$QFILE, maxambig=0, maxhomop=8, qwindowaverage=35, qwindowsize=50, processors=8)"  
done

@@

mnt/EXT/Schloss-data/bin/mothur "#unique.seqs(fasta=combined.trim.fasta)"

@@

```{r, engine='bash', eval=FALSE}
mnt/EXT/Schloss-data/bin/mothur Escobar.batch
```

@@

Now this data set included the sff files so we can use sffinfo function and start from the very start of the mothur 454 pipleine.  However, to keep things consistent we will keep with the fastq approach that we have been using from the previous data sets.  The below program was used and you can use either *#!/bin/sh* or *#!/bin/bash* and text editor to create it and then run it.

@@

# Make a directory to store the reference files
# Remember to move these back into the necessary directory later
# When you get to the batch file point,
mkdir reference.files
mv silva.bacteria.fasta reference.files
mv trainset9_032012.pds.tax reference.files
mv trainset9_032012.pds.fasta reference.files

@@

for sample2 in $(ls *fastq)
do
	mnt/EXT/Schloss-data/bin/mothur "#fastq.info(fastq=$sample2)"	
	mv $sample2 completed.fastq
	mv *logfile logfile.completed

@@

# Trim the fastas according to quality 
mkdir first.trim.completed
for Trimming in $(ls *fasta)
do
	Quality=${Trimming//fasta/qual}
	mnt/EXT/Schloss-data/bin/mothur "#trim.seqs(fasta=$Trimming, qfile=$Quality, maxambig=0, maxhomop=8, qwindowaverage=35, qwindowsize=50)"
	mv $Trimming first.trim.completed
	mv $Quality first.trim.completed
	mv *logfile logfile.completed
done

@@

```{r, engine='bash', eval=FALSE}
mnt/EXT/Schloss-data/bin/mothur Zupancic.batch
```

With that you should now be able to generate the subsample shared file that was used for the downstream analysis.
<br><br>

##### **Human Microbiome Project (HMP) 2011:**  
This data set consisted of healthy individuals that were sampled at multiple body sites and underwent extensive medical work up for metadata collection.  In order to be included in this study you could not have any underlying medical conditions.  The project website can be found [here](http://hmpdacc.org/).  The project accession number in the SRA is SRP002860.  The sequence processing pipeline can be found in an ipython notebook located [here](http://nbviewer.jupyter.org/gist/pschloss/9815766/notebook.ipynb).  The one exception to the workflow provided in the notebook is that you will need to make sure that you do not use a closed reference data base approach for OTU clustering and instead use an open reference approach.  We used the default approach listed rather than repeat the whole analysis again.  
<br><br>

@@

# Combine the qaulity and fasta files together
cat *fasta > combined.fasta
cat *qual > combined.qual

@@

After you have done this you can run a batch file with mothur and that should generate the subsample shared file that we used for this anlaysis.  It should be noted that this data set came with the barcodes and they needed to be removed so this differed form the previous data sets in this regard.  However, this is all taken care of in the batch file.

@@

mnt/EXT/Schloss-data/bin/mothur Wu.batch
```

##### **Turnbaugh, et al. 2008:**  
This is the very famous twin study of the microbiome and obesity.  The other two re-analysis articles [@finucane_taxonomic_2014; @walters_meta-analyses_2014] also use this data set.  The raw data is not avialable on the SRA but you can download a fasta file for either the V2 or V6 16S rRNA gene [here](https://gordonlab.wustl.edu/NatureTwins_2008/TurnbaughNature_11_30_08.html).  We used the V2 region.  For the most part it was relatively straight forward.  The code used and instruction are below:

```{r, engine='bash', eval=FALSE}
grep -e '^>' V2.fasta > test.fasta  
sed 's/>//' test.fasta > test2.fasta    
sed 's/\s.*$//' test2.fasta > test3.fasta  
paste -d" " test2.fasta test3.fasta > V2.groups''
```

Once we have a groups file we can just create a batch file and put it into mothur to run and come back when it is completed.
```{r, engine='bash', eval=FALSE}
mnt/EXT/Schloss-data/bin/mothur Turnbaugh.batch
```
